---
title: "Cox regression"
author: "Шавронская Дарья Олеговна"
date: "2023-01-21"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(comment = FALSE))
```

```{r message=FALSE}
library(dplyr)
library(survival)
library(data.table)
library(tibble)
library(readr)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(contsurvplot)
library(ggplot2)
library(riskRegression)
library(pammtools)
library(survminer)
```

## 1. PREPARE INITIAL DATA

1.1 clinical sample and clinical patient data
```{r}
samples <- read.delim("data_clinical_sample.txt", skip = 4)
patients <- read.delim("data_clinical_patient.txt", skip = 4)

patients <- patients %>%
  dplyr::select(PATIENT_ID, OS_STATUS, OS_MONTHS, DSS_STATUS, AGE, SEX)%>%
  filter(!(DSS_STATUS=="0:ALIVE OR DEAD TUMOR FREE" & OS_STATUS == "1:DECEASED"))%>%
  dplyr::select(!"DSS_STATUS")%>%
  na.omit()%>%
  mutate(`OS_STATUS` = ifelse(`OS_STATUS` == "1:DECEASED", 1, 0))

samples <- samples %>%
  dplyr::select(PATIENT_ID, SAMPLE_ID, SAMPLE_TYPE)
```

1.2 clinical metastasis and all clinical

```{r}
meta_clin <- inner_join(patients, samples, by = "PATIENT_ID") %>%
  filter(SAMPLE_TYPE == "Metastasis") %>%
  filter(SAMPLE_ID != "TCGA-GN-A269-01")

clin <- inner_join(patients, samples, by = "PATIENT_ID")
```

1.3 trascriptomic data

```{r}
transcriptomic <- read_tsv("data_mrna_seq_v2_rsem.txt", show_col_types = FALSE)
transcriptomic <- transcriptomic %>%
  dplyr::select(-Hugo_Symbol) %>%
  distinct(transcriptomic$Entrez_Gene_Id, .keep_all = TRUE)

colnames(transcriptomic)[1] <- "ENTREZID"

```

1.4 change entrez id to official symbols

```{r message=FALSE}
annots <- AnnotationDbi::select(org.Hs.eg.db, keys=as.character(transcriptomic$ENTREZID), columns=c("SYMBOL"), keytype="ENTREZID")

annots$ENTREZID <- as.double(annots$ENTREZID)  

RNA_symb <- inner_join(annots,transcriptomic, by = "ENTREZID")

RNA_symb <- RNA_symb %>%
  dplyr::select(!"ENTREZID")
  

```

1.5 transpose, log and scaling 

```{r}
RNA <- transpose(RNA_symb[-1])

colnames(RNA) <- RNA_symb$SYMBOL

rownames(RNA) <- colnames(RNA_symb)[-1]

RNA <- log2(RNA[-1] + 1)

RNA <- apply(RNA, 2, function(x) (x - median(x))/mad(x))

RNA <- as.data.frame(RNA)

RNA <- rownames_to_column(RNA, "SAMPLE_ID")

```

1.6 collect data 

```{r}
RNA_meta <- inner_join(meta_clin,RNA, by = "SAMPLE_ID")

RNA_clin <- inner_join(clin, RNA, by = "SAMPLE_ID")
```


## 2 UNIVARIATE ANALYSIS OF OS

### molecular signatures from articles

founded

* SUCO - 51430 
* BTN3A1 - 11119
* GZMB - 3002
* C1QA - 712
* C1QB - 713
* LINC02908 - C9orf139 - 401563
* MIR3667HG - C22orf34 - 348645
* CCL4 - 6351
* CXCL10 - 3627
* CCL5 - 6352

not founded

* MIR17 - hsa-miR-17-5p - 406952
* HLA-DQB1-AS1 - 106480429

### molecular signatures from Olga's work

founded 

* KLRD1 - 3824
* GBP4 - 115361
* OCA2 - 4948
* CCL8 - 6355
* GBP1P1 - 400759
* HSPA7 - 3311
* WNK2 - 65268



2.1 create df for a forest plot

2.1.1 articles

```{r}
covariates1 <- c("SUCO", "BTN3A1", "GZMB", "C1QA", "C1QB", "LINC02908", "MIR3667HG", "CCL4", "CXCL10", "CCL5")
univ_formulas1 <- sapply(covariates1,
                        function(x) as.formula(paste('Surv(OS_MONTHS,OS_STATUS)~', x)))
                        
univ_models1 <- lapply(univ_formulas1, function(x){coxph(x, data = RNA_meta)})
univ_results1 <- lapply(univ_models1,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=4)
                          wald.test<-signif(x$wald["test"], digits=4)
                          beta<-signif(x$coef[1], digits=4)
                          HR <-signif(x$coef[2], digits=4)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 4)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],4)
                          res<-c(beta, HR, HR.confint.lower, HR.confint.upper, wald.test, p.value)
                          
                          names(res)<-c("B values", "HR", "lower", "upper","wald.test", 
                                        "p.value")
                          return(res)
                         })
res1 <- t(as.data.frame(univ_results1, check.names = FALSE))
df1 <- as.data.frame(res1)
df1 <- rownames_to_column(df1, "MS")
```

2.2.2. Olga's work

```{r}
covariates2 <- c("KLRD1", "GBP4", "OCA2", "CCL8", "GBP1P1", "HSPA7", "WNK2")
univ_formulas2 <- sapply(covariates2,
                        function(x) as.formula(paste('Surv(OS_MONTHS,OS_STATUS)~', x)))
                        
univ_models2 <- lapply(univ_formulas2, function(x){coxph(x, data = RNA_meta)})
univ_results2 <- lapply(univ_models2,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=4)
                          wald.test<-signif(x$wald["test"], digits=4)
                          beta<-signif(x$coef[1], digits=4)
                          HR <-signif(x$coef[2], digits=4)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 4)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],4)
                          res<-c(beta, HR, HR.confint.lower, HR.confint.upper, wald.test, p.value)
                          
                          names(res)<-c("B values", "HR", "lower", "upper","wald.test", 
                                        "p.value")
                          return(res)
                         })
res2 <- t(as.data.frame(univ_results2, check.names = FALSE))
df2 <- as.data.frame(res2)
df2 <- rownames_to_column(df2, "MS")
```


2.2 create df for a comparative table

2.2.1 articles

```{r}
univ_formulas3 <- sapply(covariates1,
                        function(x) as.formula(paste('Surv(OS_MONTHS,OS_STATUS)~', x)))
                        
univ_models3 <- lapply(univ_formulas3, function(x){coxph(x, data = RNA_meta)})
univ_results3 <- lapply(univ_models3,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=4)
                          wald.test<-signif(x$wald["test"], digits=4)
                          beta<-signif(x$coef[1], digits=4)
                          HR <-signif(x$coef[2], digits=4)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 4)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],4)
                          HR <- paste0(HR, " (", HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          
                          names(res)<-c("B values", "HR (95% CI)","wald.test", 
                                        "p.value")
                          return(res)
                         })

res3 <- t(as.data.frame(univ_results3, check.names = FALSE))
table3 <- as.data.frame(res3)
table3 <- rownames_to_column(table3, "MS")
table3
```

2.2.2 Olga's work

```{r}
univ_formulas4 <- sapply(covariates2,
                        function(x) as.formula(paste('Surv(OS_MONTHS,OS_STATUS)~', x)))
univ_models4 <- lapply(univ_formulas4, function(x){coxph(x, data = RNA_meta)})
univ_results4 <- lapply(univ_models4,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=4)
                          wald.test<-signif(x$wald["test"], digits=4)
                          beta<-signif(x$coef[1], digits=4)
                          HR <-signif(x$coef[2], digits=4)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 4)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],4)
                          HR <- paste0(HR, " (", HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          
                          names(res)<-c("B values", "HR (95% CI)","wald.test", 
                                        "p.value")
                          return(res)
                         })

res4 <- t(as.data.frame(univ_results4, check.names = FALSE))
table4 <- as.data.frame(res4)
table4 <- rownames_to_column(table4, "MS")
table4
```


2.3 forest plot

2.3.1. articles

```{r}
ggplot(data=df1, aes(y=MS, x=HR, xmin=lower, xmax=upper)) +
  geom_point() + 
  geom_errorbarh(height=.1) +
  geom_vline(xintercept=1, color='black', linetype='dashed', alpha=.5) +
  labs(x='HR', y = 'Molecular signatures') +
  theme_minimal()
```

2.3.2. Olga's work

```{r}
ggplot(data=df2, aes(y=MS, x=HR, xmin=lower, xmax=upper)) +
  geom_point() + 
  geom_errorbarh(height=.1) +
  geom_vline(xintercept=1, color='black', linetype='dashed', alpha=.5) +
  labs(x='HR', y = 'Molecular signatures') +
  theme_minimal()
```

2.4 Kaplan-Meier plot

```{r}
survplotdata <- RNA_meta[,c("OS_MONTHS", "OS_STATUS", "SUCO", "BTN3A1", "GZMB", "C1QA", "C1QB", "LINC02908", "MIR3667HG", "CCL4", "CXCL10", "CCL5", "KLRD1", "GBP4", "OCA2", "CCL8", "GBP1P1", "HSPA7", "WNK2")] 
survplotdata <- as.data.frame(survplotdata)
head(survplotdata)
```
2.4.1 articles

```{r}
Expr <- 1
for (x in c("SUCO", "BTN3A1", "GZMB", "C1QA", "C1QB", "LINC02908", "MIR3667HG", "CCL4", "CXCL10", "CCL5"))
{
  survplotdata[[x]] <- ifelse(survplotdata[[x]] >= Expr, 'High', 'Low')
  survplotdata[[x]] <- factor(survplotdata[[x]],
    levels = c('Low', 'High'))
}


for (x in c("SUCO", "BTN3A1","GZMB", "C1QA", "C1QB", "LINC02908", "MIR3667HG", "CCL4", "CXCL10", "CCL5"))
{
plot <- ggsurvplot(survfit(Surv(OS_MONTHS, OS_STATUS) ~ survplotdata[[x]], 
          data = survplotdata), 
          data = survplotdata, 
          break.time.by = 100,
          pval = TRUE, 
          legend.labs = c("Low", "High"), 
          ggtheme = theme_minimal(),
          xlim = c(0,350), 
          conf.int = TRUE, 
          risk.table = TRUE, 
          legend.title = x)
print(plot)
}


```


2.4.2 Olga's work


```{r}
Expr <- 1
for (x in c("KLRD1", "GBP4", "OCA2", "CCL8", "GBP1P1", "HSPA7", "WNK2"))
{
  survplotdata[[x]] <- ifelse(survplotdata[[x]] >= Expr, 'High', 'Low')
  survplotdata[[x]] <- factor(survplotdata[[x]],
    levels = c('Low', 'High'))
}


for (x in c("KLRD1", "GBP4", "OCA2", "CCL8", "GBP1P1", "HSPA7", "WNK2"))
{
plot <- ggsurvplot(survfit(Surv(OS_MONTHS, OS_STATUS) ~ survplotdata[[x]], 
          data = survplotdata), 
          data = survplotdata, 
          break.time.by = 100,
          pval = TRUE, 
          legend.labs = c("Low", "High"), 
          ggtheme = theme_minimal(),
          xlim = c(0,350), 
          conf.int = TRUE, 
          risk.table = TRUE, 
          legend.title = x)
print(plot)
}

```
+ plot_surv_contour (Olga's work)


```{r}
survplotdata <- RNA_meta[,c("OS_MONTHS", "OS_STATUS", "SUCO", "BTN3A1", "GZMB", "C1QA", "C1QB", "LINC02908", "MIR3667HG", "CCL4", "CXCL10", "CCL5", "KLRD1", "GBP4", "OCA2", "CCL8", "GBP1P1", "HSPA7", "WNK2")] 
survplotdata <- as.data.frame(survplotdata)
survplotdata <- as.data.frame(apply(survplotdata, 2, as.numeric))

```

```{r}
model <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ GBP4, data = survplotdata, x=TRUE)
plot_surv_contour(model = model, 
                          time="OS_MONTHS",
                          status="OS_STATUS",
                          variable="GBP4",
                          data=survplotdata,
                          legend.title="GBP4",
                           bins=5)
```

```{r}
model <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ KLRD1, data = survplotdata, x=TRUE)
plot_surv_contour(model = model, 
                          time="OS_MONTHS",
                          status="OS_STATUS",
                          variable="KLRD1",
                          data=survplotdata,
                          legend.title="KLRD1",
                          bins=5)
```
```{r}
model <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ OCA2, data = survplotdata, x=TRUE)
plot_surv_contour(model = model, 
                          time="OS_MONTHS",
                          status="OS_STATUS",
                          variable="OCA2",
                          data=survplotdata,
                          legend.title="OCA2",
                          bins=5)
```

```{r}
model <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ CCL8, data = survplotdata, x=TRUE)
plot_surv_contour(model = model, 
                          time="OS_MONTHS",
                          status="OS_STATUS",
                          variable="CCL8",
                          data=survplotdata,
                          legend.title="CCL8",
                          bins=5)
```
```{r}
model <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ GBP1P1, data = survplotdata, x=TRUE)
plot_surv_contour(model = model, 
                          time="OS_MONTHS",
                          status="OS_STATUS",
                          variable="GBP1P1",
                          data=survplotdata,
                          legend.title="GBP1P1",
                          bins=5)
```

```{r}
model <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ HSPA7, data = survplotdata, x=TRUE)
plot_surv_contour(model = model, 
                          time="OS_MONTHS",
                          status="OS_STATUS",
                          variable="HSPA7",
                          data=survplotdata,
                          legend.title="HSPA7",
                          bins=5)
```

```{r}
model <- coxph(Surv(OS_MONTHS, OS_STATUS) ~ WNK2, data = survplotdata, x=TRUE)
plot_surv_contour(model = model, 
                          time="OS_MONTHS",
                          status="OS_STATUS",
                          variable="WNK2",
                          data=survplotdata,
                          legend.title="WNK2",
                          bins=5)
```

